---
title: "Iteration and List Columns"
output:
  html_document: 
    toc: true
    toc_float: true
---

We’ve noted that functions are helpful when you repeat code more than twice; we’ve also noted that a lot of statistical methods involve doing the same thing a large number of times. Putting those together motivates a careful approach to iteratation.

Meanwhile, R's data structures, especially data frames, are surprisingly flexible. This is useful when the "observations" you want to store become more complex than single values; for example, each row many contain a few scalar observations as well a complete data set. In these cases, _list columns_ are an appropriate column type, and `map` functions provide a way to interact with those columns.

This is the second module in the [Iteration](topic_iteration.html) topic; the relevant slack channel is [here](https://p8105fall2019.slack.com/messages/CN8KK99NJ).

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

library(tidyverse)

theme_set(theme_bw() + theme(legend.position = "bottom"))
```


## Some slides

<script async class="speakerdeck-embed" data-id="f0d309fcbf2c4d5d9d121eee4a6144ee" data-ratio="1.77777777777778" src="//speakerdeck.com/assets/embed.js"></script>
<div style="margin-bottom:5px"> <strong> <a href="https://speakerdeck.com/jeffgoldsmith/p8105-iteration-and-listcols" title="Iteration and List Columns" target="_blank">Iteration and List Columns</a> </strong> from <strong><a href="https://speakerdeck.com/jeffgoldsmith" target="_blank">Jeff Goldsmith</a></strong>. </div><br>


## Example

I'll write code for today's content in a new R Markdown document called `iteration_and_listcols.Rmd` in the `iteration` repo / directory I started last time. The code chunk below loads the `tidyverse` and sets a seed for reproducibility.

```{r}
library(tidyverse)

set.seed(1)
```

Things are gonna get a little weird...


### Lists

We need a brief digression about lists before we do anything. 

In R, vectors are limited to a single data class -- all elements are characters, or all numeric, or all logical. Trying to join the following vectors will result in **coersion**, as would creating vectors of mixed types.

```{r}
vec_numeric = 5:8
vec_char = c("My", "name", "is", "Jeff")
vec_logical = c(TRUE, TRUE, TRUE, FALSE)
```

Lists provide a way to store anything you want. This flexibility is great, but is offset by a certain ... clunkiness. Lists contain indexed elements, and the indexed elements themselves be scalars, vectors, or other things entirely. 

```{r}
l = list(vec_numeric = 5:8,
         mat         = matrix(1:8, 2, 4),
         vec_logical = c(TRUE, FALSE),
         summary     = summary(rnorm(1000)))
l
```

Lists can be accessed using names or indices, and the things in lists can be accessed in the way you would usually access an object of that type. 

```{r}
l$vec_numeric
l[[1]]
l[[1]][1:3]
```

Lists seem bizarre but are really useful. Right now, we'll use them to hold general inputs and outputs of iterative processes. Even more importantly, we'll see that data frames are actually a very specific kind of list -- one comprised of vectors of the same length -- which is why they can store variables of different types. 


### `for` loops

For this example, I'm going to start with the pretty simple data frame defined below, and confirm that "under the hood" this is a list.

```{r}
df = tibble(
  a = rnorm(20, 3, 1),
  b = rnorm(20, 0, 5),
  c = rnorm(20, 10, .2),
  d = rnorm(20, -3, 1)
)

is.list(df)
```

I'd like to apply my simple `mean_and_sd` function from [writing functions](writing_functions.html) to each column of this dataframe. For completeness, that function is below. 

```{r}
mean_and_sd = function(x) {
  
  if (!is.numeric(x)) {
    stop("Argument x should be numeric")
  } else if (length(x) == 1) {
    stop("Cannot be computed for length 1 vectors")
  }
  
  mean_x = mean(x)
  sd_x = sd(x)

  tibble(
    mean = mean_x, 
    sd = sd_x
  )
}
```

We can apply the `mean_and_sd` function to each column of `df` using the lines below. Throughout this content, I'll take advantage of the fact that data frames are a kind of list -- keeping this in mind when you're iterating is _really_ useful .

```{r}
mean_and_sd(df[[1]])
mean_and_sd(df[[2]])
mean_and_sd(df[[3]])
mean_and_sd(df[[4]])
```

But now we've broken our "don't repeat code more than twice" rule! Specifically, we've applied the same function / operation to the elements of a list sequentially. This is exactly the kind of code repetition for loops address

Below, I define an output list with the same number of entries as my target dataframe; a sequence to iterate over; and a for loop body that applies the `mean_and_sd` function for each sequence element and saves the result.

```{r}
output = vector("list", length = 4)

for (i in 1:4) {
  output[[i]] = mean_and_sd(df[[i]])
}
```

This is already much cleaner than using four almost-identical lines of code, and will make life easier the larger our sequence gets. 

In this example, I bypassed a common first step in writing loops because I already had the function I wanted to repeat. Frequently, however, I'll start with repeated code segements, then abstract the underlying process into a function, and then wrap things up in a for loop. 


### `map`

A criticism of `for` loops is that there's a lot of overhead -- you have to define your output vector / list, there's the `for` loop bookkeeping to do, etc -- that distracts from the purpose of the code. In this case, we want to apply `mean_and_sd` to each column of `df`, but we have to scan inside the for loop to figure that out.

The `map` functions in `purrr` try to make the purpose of your code clear. Compare the loop above to the line below.

```{r}
output = map(df, mean_and_sd)
```

The first argument to `map` is the vector / list ( / data frame) we want to iterate over, and the second argument is the function we want to apply to each element. The line above will produce the same output as the previous loop, but is clearer and easier to understand (once you're used to `map` ...). 

It's sometimes necessary to be more specific in giving arguments to `map`. In particular, using `.x = df` for the input list and `~ mean_and_sd(.x)` to specify the function applied to the input list (using `.x` as a placeholder) will produce the same result. 

```{r}
output = map_df(.x = df, ~ mean_and_sd(.x))
```

This code (using `map`) is why we pointed out in [writing functions](writing_functions.html) that functions can be passed as arguments to other functions. The second argument in `map(df, mean_and_sd)` is a function we just wrote. To see how powerful this can be, suppose we wanted to apply a different function, say `median`, to each column of `df`. The chunk below includes both the loop and the `map` approach.

```{r}
output = vector("list", length = 4)

for (i in 1:4) {
  output[[i]] = median(df[[i]])
}

output = map(df, median)
# output = map(.x = df, ~ median(.x))
```

Again, both options produce the same `output`, but the `map` places the focus squarely on the function you want to apply by removing much of the bookkeeping. 

### `map` variants

There are some useful variants to the basic `map` function if you know what kind of output you're going to produce. Below we use `map_dbl` because `median` outputs a single numeric value each time; the result is a vector instead of a list. Using the `.id` argument keeps the names of the elements in the input list. 

```{r}
output = map_dbl(df, median, .id = "input")
```

If we tried to use `map_int` or `map_lgl`, we'd get an error because the output of `median` isn't a integer or a logical. This is a good way to help catch mistakes when they arise.

Similarly, since we know `mean_and_sd` produces a data frame, we can use the output-specific `map_dfr`; this will produce a single data frame. 

```{r}
output = map_dfr(df, mean_and_sd, .id = "input")
```

The `map_df` variants can be helpful when your map statement is part of a longer chain of piped commands. 

Finally, `map2` (and `map2_dbl`, etc) is helpful when your function has two arguments. In these cases, I find it very helpful to be specific about arguments using something like the following:

```{r, eval = FALSE}
output = map2(.x = input_1, .y = input_2, ~func(arg_1 = .x, arg_2 = .y))
```

Although we won't go into this in detail, one example where you might use this is in loading the LoTR data using functions and inputs we defined previously.

```{r, eval = FALSE}
lotr_data = map2_df(
  .x = cell_ranges, .y = movie_names, 
  ~lotr_load_and_tidy(path = "./data/LotR_Words.xlsx", range = .x, movie_name = .y))
```


### Revisiting Napoleon

In [reading data from the web](reading_data_from_the_web.html) and elsewhere, we wrote code that allowed us to scrape information in Amazon reviews; in [writing functions](writing_functions.html) we wrapped that code into a function called `read_page_reviews` which, for a given `url`, produces a data frame containing review titles, star ratings, and text.

```{r}
library(rvest)

read_page_reviews = function(url) {
  
  h = read_html(url)
  
  title = h %>%
    html_nodes("#cm_cr-review_list .review-title") %>%
    html_text()
  
  stars = h %>%
    html_nodes("#cm_cr-review_list .review-rating") %>%
    html_text() %>%
    str_extract("\\d") %>%
    as.numeric()
  
  text = h %>%
    html_nodes(".review-data:nth-child(5)") %>%
    html_text()
  
  data_frame(title, stars, text)
}
```

**_Learning Assessment:_** Use this function to read the five pages of reviews on the URLs defined in the code chunk below.

```{r}
url_base = "https://www.amazon.com/product-reviews/B00005JNBQ/ref=cm_cr_arp_d_viewopt_rvwer?ie=UTF8&reviewerType=avp_only_reviews&sortBy=recent&pageNumber="
vec_urls = str_c(url_base, 1:5)
```

<details>
<summary> Solution </summary>

First I'll define a vector of URLs to act as an input, and then I'll iterate over that vector using both a for loop and a `map_df` statement.

```{r, eval = FALSE}
output = vector("list", 5)

for (i in 1:5) {
  output[[i]] = read_page_reviews(vec_urls[[i]])
}

dynamite_reviews = bind_rows(output)

dynamite_reviews = map_df(vec_urls, read_page_reviews)
```

As with previous examples, using a for loop is pretty okay but the `map_df` call is clearer.

</details>


### List columns

Shifting gears a bit, let's revisit the weather data from [visualization](visualization.html) and elsewhere; these data consist of one year of observations from three monitoring stations. The code below pulls these data into R (using the `rnoaa` package, which interacts with the NOAA API).

```{r weather_data_create, cache = TRUE}
weather = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USC00519397", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2016-01-01",
    date_max = "2016-12-31") %>%
  mutate(
    name = recode(id, USW00094728 = "CentralPark_NY", 
                      USC00519397 = "Waikiki_HA",
                      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) %>%
  select(name, id, everything())
```

The station name and id are constant across the year's temperature and precipitation data. For that reason, we can reorganize these data into a new data frame with a single row for each station. Weather data will be separated into three station-specific data frames, each of which is the data "observation" for the respective station.

```{r}
weather_nest = 
  nest(weather, data = date:tmin)

weather_nest
```

Here I've used `nest` by specifying a column range to collapse within remaining variable values. 

The `name` column is a character column -- if you pull this column from the `weather` data frame, the result is a character vector. Similarly, the `data` column is a *list column* -- on it's own, it's a list. 

```{r}
weather_nest %>% pull(name)
weather_nest %>% pull(data)
```

The list column really is a list, and will behave as such elsewhere in R. So, for example, you can examine the first list entry using usual list index procedures.

```{r}
weather_nest$data[[1]]
```

Of course, if you can `nest` data you should be able to `unnest` it as well, and you can (with the caveat that you're unnesting a list column that contains a data frame).

```{r}
unnest(weather_nest, cols = data)
```

Nesting columns can help with data organization and comprehension by masking complexity you're less concerned about right now and clarifying the things you are concerned about. In the weather data, it can be helpful to think of stations as the basic unit of observation, and daily weather recordings as a more granular level of observation. Nesting can also simplify the use of analytic approaches across levels of a higher variable. 

### Operations on list columns

You will need to be able to manipulate list columns, but usual operations for columns that might appear in `mutate` (like `mean` or `recode`) often don't apply to the entries in a list column. 

Instead, recognizing list columns as **_list_** columns motivates an approach for working with nested data frames.

Suppose we want to fit the simple linear regression relating `tmax` to `tmin` for each station-specific data frame. First I'll write a quick function that takes a data frame as the sole argument to fit this model.

```{r}
weather_lm = function(df) {
  lm(tmax ~ tmin, data = df)
}
```

Let's make sure this works on a single data frame.

```{r}
weather_lm(weather_nest$data[[1]])
```

Great! Keeping in mind that `weather$data` is a **_list_**, we can apply our `weather_lm` function to each data frame using `map`. 

```{r}
map(weather_nest$data, weather_lm)
```

You can avoid the creation of a dedicated function using `map`'s syntax for "anonymous" (i.e. not named and saved) functions. This is fine for really short operations, but I typically write a dedicated function instead of using this option.

```{r}
map(weather_nest$data, ~lm(tmax ~ tmin, data = .x))
```

The `map` function returns a _list_; we could store the results as a new **_list column_** ... !!!

We've been using `mutate` to define a new variable in a data frame, especially one that is a function of an existing variable. That's exactly what we will keep doing. 

```{r}
weather_nest = 
  weather_nest %>% 
  mutate(models = map(data, weather_lm))

weather_nest
```

This is great! We now have a data frame that has rows for each station; columns contain weather datasets and fitted models. This makes it very easy to keep track of models across stations, and to perform additional analyses. 

This is, for sure, a fairly complex bit of code, but in just a few lines we're able to fit separate linear models to each of our stations. And, once you get used to list columns, `map`, and the rest of it, these lines of code are pretty clear and can be extended to larger datasets with more complex structures.


### Revisiting past examples

Earlier, we looked at scraping reviews of Napoleon Dynamite from several pages on Amazon. This required the creation of input and output vectors / lists; now we'll repeat this using data frames and list columns.

```{r}
dynamite_reviews = 
  tibble(page = 1:5,
         urls = str_c(url_base, page)) %>% 
  mutate(reviews = map(urls, read_page_reviews)) %>% 
  unnest()
```

I'll also revisit a way to load and tidy the LoTR words data that starts with a dataframe containing cell ranges for each movie. Of all the approaches we've seen, this is probably my favorite.

```{r}
lotr_cell_ranges = 
  tibble(
    movie = c("fellowship_ring", "two_towers", "return_king"),
    cells = c("B3:D6", "F3:H6", "J3:L6")
  )

lotr_tidy = 
  lotr_cell_ranges %>% 
  mutate(
    word_data = map(cells, ~readxl::read_excel("./data/LotR_Words.xlsx", range = .x))
  ) %>% 
  unnest(cols = word_data) %>% 
  janitor::clean_names() %>% 
  pivot_longer(
    female:male,
    names_to = "sex",
    values_to = "words") %>%
  mutate(race = str_to_lower(race)) %>% 
  select(movie, everything(), -cells) 
```



## Other materials

Iteration can be tricky -- the readings below will help as you work through this!

* The chapter on [iteration](http://r4ds.had.co.nz/iteration.html) in R for Data Science has a useful treatment of iteration using loops and `map`
* Jenny Bryan's [`purrr` tutorial](https://jennybc.github.io/purrr-tutorial/) has a lot of useful information and examples
* R Programming for Data Science has information on [loops](https://bookdown.org/rdpeng/rprogdatascience/control-structures.html#for-loops) and [loop functions](https://bookdown.org/rdpeng/rprogdatascience/loop-functions.html); given Roger Peng's tendency towards base R he focuses on `lapply` and others instead of `map`
* This [question and response](https://stackoverflow.com/questions/45101045/why-use-purrrmap-instead-of-lapply) on stack overflow explains why one might prefer `map` to `lapply`

List columns take some getting used to; there are some materials to help with that.

* R for Data Science has a chapter on [fitting many models](http://r4ds.had.co.nz/many-models.html)
* Jenny Bryan's [purrr tutorial](https://jennybc.github.io/purrr-tutorial/) has useful list-column examples
* The [R file](https://github.com/tidyverse/dplyr/blob/master/data-raw/starwars.R) used to create the `starwars` dataset (in the `tidyverse`) using the [Star Wars API](https://swapi.co) processes list output (from the API) using several `map` variants

The code that I produced working examples in lecture is [here](https://github.com/p8105/iteration).
