---
title: "Simulation"
output:
  html_document: 
    toc: true
    toc_float: true
---

We've noted that functions are helpful when you repeat code more than twice; we've also noted that a lot of statistical methods involve doing the same thing a large number of times. Simulation is a common statistical approach that takes advantage of the ability to iterate many times using computers. 

This is the third module in the [Iteration](topic_iteration.html) topic; the relevant slack channel is [here](https://p8105fall2019.slack.com/messages/CN8KK99NJ).

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

library(tidyverse)

theme_set(theme_bw() + theme(legend.position = "bottom"))
```

## Some slides

<script async class="speakerdeck-embed" data-id="285289b17d194a4282d53f1800d37199" data-ratio="1.77777777777778" src="//speakerdeck.com/assets/embed.js"></script>
<div style="margin-bottom:5px"> <strong> <a href="https://speakerdeck.com/jeffgoldsmith/p8105-simulation" title="Simulation and Bootstrapping" target="_blank">Simulation</a> </strong> from <strong><a href="https://speakerdeck.com/jeffgoldsmith" target="_blank">Jeff Goldsmith</a></strong>. </div><br>


## Example

I'll write code for today's content in a new R Markdown document called `simulation.Rmd` in the `iteration` directory / repo. The code chunk below loads the usual packages and sets a seed for reproducibility.

```{r}
library(tidyverse)

set.seed(1)
```

### Simulation: SLR for one $n$

In [writing functions](writing_functions.html) we wrote a short function to simulate data from a simple linear regression, fit the regression model, and return estimates of regression coefficients. Specifically, we generate data from 
$$ y_i = \beta_0 + \beta_1 x_i + \epsilon_i $$

for subjects $1 \leq i \leq n$ with $\epsilon_i \sim N[0,1]$ and return estimates 

$\hat{\beta}_0, \hat{\beta}_1$. That function is below. 

```{r}
sim_regression = function(n, beta0 = 2, beta1 = 3) {
  
  sim_data = tibble(
    x = rnorm(n, mean = 1, sd = 1),
    y = beta0 + beta1 * x + rnorm(n, 0, 1)
  )
  
  ls_fit = lm(y ~ x, data = sim_data)
  
  tibble(
    beta0_hat = coef(ls_fit)[1],
    beta1_hat = coef(ls_fit)[2]
  )
}
```

Important statistical properties of estimates $\hat{\beta}_0, \hat{\beta}_1$ are established under the conceptual framework of repeated sampling. If you could draw from a population over and over, your estimates will have a known mean and variance:

$$ \hat{\beta}_0 \sim \left[\beta_0, \sigma^2 \left(\frac{1}{n} + \frac{\bar{x}}{\sum (x_i - \bar{x})^2}\right) \right] \mbox{ and } \hat{\beta}_1 \sim \left[\beta_1,\frac{\sigma^2}{\sum (x_i - \bar{x})^2} \right] $$

(Because our simulation design generates errors from a Normal distribution we also know that the estimates follow a Normal distribution, although that's not guaranteed by least squares estimation.)

In the real world, drawing samples is time consuming and costly, so "repeated sampling" remains conceptual. On a computer, though, drawing samples is pretty easy. That makes simulation an appealing way to examine the statistical properties of your estimators. 

Let's run `sim_regression()` 100 times to see the effect of randomness in $\epsilon$ on estimates $\hat{\beta}_0, \hat{\beta}_1$.

```{r loop_sim, cache = TRUE}
output = vector("list", 100)

for (i in 1:100) {
  output[[i]] = sim_regression(30)
}

sim_results = bind_rows(output)
```

Taking a look at the `for` loop we used to create these results, you might notice that there's no `input` list -- the sequence is used to keep track of the output but doesn't affect the computation performed inside the `for` loop. In cases like these, the `purrr::rerun` function is very handy. 

```{r}
sim_results = 
  rerun(100, sim_regression(30, 2, 3)) %>% 
  bind_rows()
```

Structurally, `rerun` is a lot like `map` -- the first argument defines the amount of iteration and the second argument is the function to use in each iteration step. As with `map`, we've replaced a for loop with a segment of code that makes our purpose much more transparent but both approaches give the same results. 

Let's make some quick plots and compute some summaries for our simulation results.

```{r}
sim_results %>% 
  ggplot(aes(x = beta0_hat, y = beta1_hat)) + 
  geom_point()

sim_results %>% 
  pivot_longer(
    beta0_hat:beta1_hat,
    names_to = "parameter", 
    values_to = "estimate") %>% 
  group_by(parameter) %>% 
  summarize(emp_mean = mean(estimate),
            emp_var = var(estimate)) %>% 
  knitr::kable(digits = 3)
```

This is **_great_**! We've seen how our estimates are distributed under our simulation scenario, and can compare empirical results to theoretical ones. In this way, we can build intution for fundamental statistical procedures under repeated sampling in a way that's not possible with single data sets. 


### Simulation: SLR for several $n$s

Sample size makes a huge difference on the variance of estimates in SLR (and pretty much every statistical method). Let's try to clarify that effect through simulating at a few sample sizes. 

I'll start this process with a for loop around the code I established above using `rerun` (I could start from scratch by nesting one for loop in another for loop, but let's not). 

```{r}
n_list = list("n_30"  = 30, 
              "n_60"  = 60, 
              "n_120" = 120, 
              "n_240" = 240)
output = vector("list", length = 4)

for (i in 1:4) {
  output[[i]] = rerun(100, sim_regression(n_list[[i]])) %>% 
    bind_rows
}
```

After this loop, `output` is a list of 4 data frames; each data frame contains the results of 100 simulations at different sample sizes. 

Before we spend time looking at the results of the simulation, let's recast this using list columns and `map`. I'll set up a tibble with the sample sizes I want to investigate, and then use `rerun` to perform the complete simulation for each sample size. Remember that `rerun` produces a list, so I'm going to use `bind_rows` to produce a tibble of simulation results for each sample size. Finally, I'll `unnest` the tibbles to produce a standard data frame with no list columns.

```{r}
sim_results = 
  tibble(sample_size = c(30, 60, 120, 240)) %>% 
  mutate(
    output_lists = map(.x = sample_size, ~rerun(1000, sim_regression(n = .x))),
    estimate_dfs = map(output_lists, bind_rows)) %>% 
  select(-output_lists) %>% 
  unnest(estimate_dfs)
```

Let's take a look at what we've accomplished in our simulations! First I'll take a look at the distribution of slope estimates across sample sizes. 

```{r}
sim_results %>% 
  mutate(
    sample_size = str_c("n = ", sample_size),
    sample_size = fct_inorder(sample_size)) %>% 
  ggplot(aes(x = sample_size, y = beta1_hat, fill = sample_size)) + 
  geom_violin()
```

These estimates are centered around the truth (3) for each sample size, and the width of the distribution shrinks as sample size grows. 

Next, I'll look at the bivariate distribution of intercept and slope estimates across sample sizes.

```{r}
sim_results %>% 
  mutate(
    sample_size = str_c("n = ", sample_size),
    sample_size = fct_inorder(sample_size)) %>% 
  ggplot(aes(x = beta0_hat, y = beta1_hat)) + 
  geom_point(alpha = .2) + 
  facet_grid(~sample_size)
```

The variability in the slope estimates was shown in the violin plot, but now we have a sense for the bivariate distribution of intercepts and slopes. Estimates of the intercept and slope are correlated with each other; this is expected from theoretical results describing the joint distribution of estimated regression coefficients. 

Lastly I'll look at the empirical mean and variance of these estimates.

```{r}
sim_results %>% 
  pivot_longer(
    beta0_hat:beta1_hat,
    names_to = "parameter", 
    values_to = "estimate") %>% 
  group_by(parameter, sample_size) %>% 
  summarize(emp_mean = mean(estimate),
            emp_var = var(estimate)) %>% 
  knitr::kable(digits = 3)
```

These values are consistent with the formulas presented above. This kind of check is a useful way to support derivations (although they don't serve as a formal proof in any way).




## Other materials

* Problem of small power leading to over estimation has been discussed [here](https://www.nature.com/articles/nrn3475) among other places

The code that I produced working examples in lecture is [here](https://github.com/p8105/iteration).
