---
title: "Statistical Learning"
output:
  html_document: 
    toc: true
    toc_float: true
---

Stat learning stuff ...

This page is related to content in [plotly and dashboards](plotly_and_dashboards.html), and the slack channel for extra topics is [here](https://p8105fall2019.slack.com/messages/CN01H0GHX).

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

library(tidyverse)

theme_set(theme_bw() + theme(legend.position = "bottom"))
```


## Some slides

<script async class="speakerdeck-embed" data-id="911acc341b434b02b24df1fdab954c7e" data-ratio="1.77777777777778" src="//speakerdeck.com/assets/embed.js"></script>
<div style="margin-bottom:5px"> <strong> <a href="https://speakerdeck.com/jeffgoldsmith/p8105-statistical-learning" title="Statistical Learning" target="_blank">Shiny</a> </strong> from <strong><a href="https://speakerdeck.com/jeffgoldsmith" target="_blank">Jeff Goldsmith</a></strong>. </div><br>


## Example

As always, I'll work on today's example in a GitHub repo + local directory / R Project. [This zip file](resources/extra_topic_data.zip) has a couple of datasets we'll look at.

```{r}
library(tidyverse)
library(glmnet)

set.seed(11)
```

### Lasso

Birthweight subset

```{r import}
bwt_df = 
  read_csv("data/birthweight.csv") %>% 
  janitor::clean_names() %>%
  mutate(
    babysex = as.factor(babysex),
    babysex = fct_recode(babysex, "male" = "1", "female" = "2"),
    frace = as.factor(frace),
    frace = fct_recode(frace, "white" = "1", "black" = "2", "asian" = "3", 
                       "puerto rican" = "4", "other" = "8"),
    malform = as.logical(malform),
    mrace = as.factor(mrace),
    mrace = fct_recode(mrace, "white" = "1", "black" = "2", "asian" = "3", 
                       "puerto rican" = "4")) %>% 
  sample_n(200)
  
x = model.matrix(bwt ~ ., bwt_df)[,-1]
y = bwt_df$bwt
```

We fit the lasso model for each tuning parameter in a pre-defined grid, and then compare the fits using cross validation.

```{r}
lambda = 10^(seq(3, -2, -0.1))

fit.glmnet =
  glmnet(x, y, lambda = lambda)

out.cv.glmnet =
  cv.glmnet(x, y, lambda = lambda)

lambda_opt = out.cv.glmnet$lambda.min
```

The plot below shows coefficient estimates corresponding to a subset of the genes in the dataset -- these are genes that have non-zero coefficients for at least one tuning parameter value in the pre-defined grid. As lambda increases, the coefficient values are shrunk to zero and the model becomes more sparse. The optimal tuning parameter, determined using cross validation, is shown by a vertical blue line.

```{r}
broom::tidy(fit.glmnet) %>% 
  select(term, lambda, estimate) %>% 
  complete(term, lambda, fill = list(estimate = 0) ) %>% 
  filter(term != "(Intercept)") %>% 
  ggplot(aes(x = log(lambda, 10), y = estimate, group = term, color = term)) + 
  geom_path() + 
  geom_vline(xintercept = log(lambda_opt, 10), color = "blue", size = 1.2) +
  theme(legend.position = "none")
```

CV curve.

```{r}
broom::tidy(out.cv.glmnet) %>% 
  ggplot(aes(x = log(lambda, 10), y = estimate)) + 
  geom_point()  
```





On the full dataset, lasso doesn't do you so much good.

### Clustering: pokemon

```{r}
poke_df = 
  read_csv("./data/pokemon.csv") %>% 
  select(HP, Speed)

poke_df %>% 
  ggplot(aes(x = HP, y = Speed)) + 
  geom_point()

kmeans_fit =
  kmeans(x = poke_df, centers = 3)

poke_df =
  broom::augment(kmeans_fit, poke_df)

poke_df %>% 
  ggplot(aes(x=HP, y=Speed, color=.cluster)) +
  geom_point()
```

```{r}
clusts =
  tibble(k = 2:4) %>%
  mutate(
    km_fit =    map(k, ~kmeans(poke_df, .x)),
    augmented = map(km_fit, ~broom::augment(.x, poke_df))
  )

clusts %>% 
  select(-km_fit) %>% 
  unnest(augmented) %>% 
  ggplot(aes(HP, Speed, color = .cluster)) +
  geom_point(aes(color = .cluster)) +
  facet_grid(~k)
```

### Clustering: trajectories

```{r}
traj_data = 
  read_csv("~/Desktop/trajectories.csv")

traj_data %>% 
  ggplot(aes(x = week, y = value, group = subj)) + 
  geom_point() + 
  geom_path()
```

```{r}
int_slope_df = 
  traj_data %>% 
  nest(data = week:value) %>% 
  mutate(
    models = map(data, ~lm(value ~ week, data = .x)),
    result = map(models, broom::tidy)
  ) %>% 
  select(subj, result) %>% 
  unnest(result) %>% 
  select(subj, term, estimate) %>% 
  pivot_wider(
    names_from = term,
    values_from = estimate
  ) %>% 
  rename(int = "(Intercept)", slope = week)

int_slope_df %>% 
  ggplot(aes(x = int, y = slope)) + 
  geom_point()

km_fit = 
  kmeans(
    x = int_slope_df %>% select(-subj) %>% scale, 
    centers = 2)

int_slope_df =
  broom::augment(km_fit, int_slope_df)

int_slope_df %>% 
  ggplot(aes(x = int, y = slope, color = .cluster)) +
  geom_point()

left_join(traj_data, int_slope_df) %>% 
  ggplot(aes(x = week, y = value, group = subj, color = .cluster)) + 
  geom_point() + 
  geom_path() 
```


This is "trajectory analysis" ...



## Other materials

* ISLR, chapters ... and ... 
* https://shiny.rstudio.com/gallery/kmeans-example.html
* https://idc9.github.io/stor390/notes/clustering/clustering.html
* https://cran.microsoft.com//web/packages/broom/vignettes/kmeans.html

The code that I produced working examples in lecture is [here](https://github.com/P8105/extra_topics).

